---
title: "MovieLens Project"
author: "Brendan Ball"
date: "7/9/2021"
output: pdf_document
---

# 1. Introduction

The MovieLens Dataset, published by F. Maxwell Harper and Joseph A. Konstan, is 
a compiled data that has 10 million ratings, 100,000 tag applications in
relation to 10,000 movies, and reviewed by 72,000 users [1]. The data, 
published in 2009 will help provide the information necessary to create a 
recommendation system using the tools learned the HarvardX: Data Science course.
This is the goal of the project, and if achieved, can help improve the user
experience when selecting a movie from recommendation sources.

The key steps that will be performed includes various features, such as genres, 
movie ratings, and popular movie titles will to be investigated. This will help
provide an informal understanding of the entire data set. Once this has been 
done, a root mean square error (RMSE) calculation can be calculated to study 
the accuracy and training of the algorithm. 

Coding techniques such as data-wrangling, data visualization, plot generation, 
and other essential topics will be utilized to analyze the data.

To first operate the R code, the following packages will need to be installed:
1) tidyverse
2) caret
3) data.table
4) kableExtra
5) lubridate

```{r Installation of Packages, include=FALSE}
# Installation of packages if needed
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", 
                                     repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(tinytex)) install.packages("tinytext", repos = "http://cran.us.r-project.org")
```

The library packages will then be uploaded into the R code after installation.

```{r Library Packages}
# Use the library commands
library(tidyverse)
library(caret)
library(data.table)
library(kableExtra)
library(lubridate)
library(tinytex)
```

The movie set is imported from the online website shown below. A 62.5MB file
is expected to be downloaded. Please ensure there is enough room to download
this large file before running the code.

```{r Downloading the Movie Data Set}
# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

# Download the temporary file (If no file is present)
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
```

The movies data that is downloadeed from the internet is manipulated to become
a clean data set. Using the following code below:

```{r Creating the Movies Data Frame Temporarily}
ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

# Create a movies dataframe
movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), 
                          "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
```

```{r Run for R 4.0 or Later}
# If using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")
```

The validation set for the model is created below:

```{r Validation Set Creation}
# Validation set will be 10% of MovieLens data

# If using R 3.5 or earlier, use `set.seed(1)`
suppressWarnings(set.seed(1, sample.kind="Rounding")) 

test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

# 2. Methods & Analysis

To better understand the current data set, a few preliminary commands are used.
The first is the class() command to see the class of the edx data.

```{r Class of the Data}
# Determine the class of the edx data
class(edx)
```

The class of edx is a data.frame

The glimpse of the edx data can also be studied. This will show all 
observations (9,000,055) and variables (6). 

```{r Glimpse of the Data Set}
# Glimpse the data set of "edx"
glimpse(edx)
```

The summary of the results is shown below. Where different information for each 
of the variables is listed.

```{r Result Summaries}
# Inspect the result summaries of the results
summary(edx)
```

After generating plots and learning more about the data, the next step to test
the performance of the model is with root mean squared error (RMSE). 

To calculate the RMSE value, residuals must be defined. This is the difference 
between the predicted value, and the actual value in a data set. In the 
following equation, $\hat{y}_{i,o}$ is denoted as the observed value, whereas $\hat{y}_{u,a}$ is the actual value. Finding the difference, and squaring this 
term as the sum of all 'u' terms, and dividing by N number of non-missing data 
points will return the RMSE value. A higher RMSE value will show as a better fit
to the data, however; we want to ensure that the model is not overtrained, 
as this will not be fully reliable in any future movies that may be released.

$$RMSE = \sqrt{(1/N)\sum_{u,i}(\hat{y}_{u,o}-\hat{y}_{u,a})^2}$$

# 3. Results

To understand the results, a genre_order data is made where the genres are 
ranked in descending order. This will prepare the data for the following table.

```{r Individual Genres}
# Explore top individual genres 
genre_order <- edx %>% separate_rows(genres, sep = "\\|") %>%
  group_by(genres) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```

The table of the top genres is generated below. The top three genres include
drama, comedy, and action in respective order. It is clear that the top three
genres are significantly more popular than the following genres. This suggests 
that the bulk of genres that are enjoyed can be grouped into drama, comedy, and
action. Perhaps thrillers and adventure movies can be ranked high as well.

```{r Table of Top Genres}
# Create a table to list the top genres and the 
table_genre <- data.table(genre_order, rownames = FALSE, filter="top")
table_genre <- table_genre[,c("genres","count")]  
table_genre
```

The top 10 movie titles are ranked in descending order. 

```{r The Top 10 Movie Titles}
# Rank the Top 10 Movie Titles
ranked_titles <- edx %>%
  group_by(title) %>%
  summarize(count=n()) %>%
  top_n(10,count) %>%
  arrange(desc(count))
```

Another visual to compare the top 10 movies based on ratings is shown below. 
Movies in the late 1990s seemed to be the most popular, with Pulp Fiction (1994)
ranking the highest rated movie, following Forrest Gump (1994) in second, 
and The Silence of the Lambs (1991) ranked third. 

```{r Bar Plot Visual Representation of Ranked Titles}
# Plot a bar graph of the top ranked titles
ranked_titles %>% 
  ggplot(aes(x=reorder(title, count), y=count)) +
  geom_bar(stat='identity', fill="dark green", color = "black") + 
  coord_flip(y=c(0, 35000)) +
  labs(x="", y="Number of ratings") +
  geom_text(aes(label= count), hjust=-0.2, size=3) +
  labs(title="Top 10 Movies Based on Ratings" , caption = "[1]")
```

Based from the earlier two figures, it can be hypothesized that the Pulp
Fiction movie could be a movie with a genre ranked in the top three or five. 
It is, in fact, confirmed when viewing the table below. Pulp Fiction is comedy, 
crime, and drama movie, with both comedy and drama ranking in the top three 
genres. The majority of the top 10 titles also show this similar pattern, with
a few exceptions to this pattern. 

```{r Output of the Top 10 Titles, message = FALSE}
# Output the Top 10 Titles, and its Corresponding Titles and Count
kable(head(edx %>% group_by(title,genres) %>%
             summarize(count = n()) %>%
             top_n(10,count) %>%
             arrange(desc(count)) ,
           10)) %>%
  kable_styling(bootstrap_options = "basic", full_width = NULL , 
                position ="center") %>%
  column_spec(1, bold = T) %>%
  column_spec(2) %>%
  column_spec(3)
```

Based on the histogram below, it seems to indicate that some users had a wider 
range of movie ratings than other users. Based on this figure, some people 
rated significantly more movies than other people. If ratings by each individual
were to be somewhat equal, we would expect a more linearly-based pattern.

```{r Movies Rated by UserId}
# Number of User Ratings by UserId
edx %>% group_by(movieId) %>% summarize(n = n()) %>%
  ggplot(aes(n)) + geom_histogram(fill = "orange", color = "black", bins = 10) +
  scale_x_log10() +
  ggtitle("Number of Movies Ratings by UserId") +
  xlab("SemilogX of User Id") +
  ylab("Number of Ratings")
```

Another question asks the way a person rates a movie. Are whole star ratings 
more common than half stars? And if so, which rating value (out of five stars) 
is the most popular selection? Below, the stars grouping is prepared for 
plotting.

```{r Preparation of the Whole/Half Star Plot}
# Generate a group between whole star and half star results
stars_grouping <-  ifelse((edx$rating == 1 |edx$rating == 2 | edx$rating == 3 | 
                  edx$rating == 4 | edx$rating == 5) ,
                   "whole star", 
                   "half star") 

# Create a data frame to use for the plot generation
stars_ratings <- data.frame(edx$rating, stars_grouping)
```

The histogram is plotted below, with 4.0 stars being the most popular choice 
among users who rate the movies. Following 4.0 stars are 3.0 stars, then 5.0 
stars, then two and one stars. No person rated 0 stars on any movie.

```{r Histogram of Whole/Half Stars}
# Generating a Histogram comparing the distribution
ggplot(stars_ratings, aes(x= edx.rating, fill = stars_grouping)) +
  geom_histogram( binwidth = 0.5, color = "black") +
  scale_x_continuous(breaks=seq(0, 5, by= 0.5)) +
  scale_fill_manual(values = c("half star"="blue", "whole star"="orange")) +
  labs(x = "Rating (Out of 5)", y = "Number of Ratings (count)", 
       caption = '[1]') +
  ggtitle("Distribution of Whole and Half Star Ratings")
```

In both the whole star and half star patterns, a similar trend follows.

Whole Stars
Highest Rating: 4 > 3 > 5 > 2 > 1 :Lowest Rating
Half Stars
Highest Rating: 3.5 > 4.5 > 2.5 > 1.5 > 1.0 :Lowest Rating

To ensure there are no significant changes in rating patterns throughout the
years, a time stamp plot is made, following the trends of ratings each year.
Based on the results below, it seems that all ratings fell within the 3-4 star
threshold. The average stayed stagnant around 3.5 stars, showing that there 
is little to no effects on the ratings based on time. This assumption will help
simplify the understanding of the data significantly.

```{r Time Stamp of the Ratings by Year}
# Generate a plot of the timestamp, with x axis being year, y being ratings
edx %>% 
  mutate(date = round_date(as_datetime(timestamp), unit = "week")) %>%
  group_by(date) %>%
  summarize(rating = mean(rating)) %>%
  ggplot(aes(date, rating)) +
  geom_point() +
  geom_smooth(method = 'loess', formula = y ~ x) +
  ggtitle("Time Stamp of the Ratings by Year")+
  labs(x = "Year", y = "Ratings (Out of 5)", caption = "[1]") +
  ylim(1, 5) + 
  geom_hline(yintercept=3, linetype="dashed", 
                color = "red", size=1) +
  geom_hline(yintercept=4, linetype="dashed", 
                color = "red", size=1)
```

The root mean squared error (RMSE) calculation is next conducted. Based on the 
plot, the RMSE increases with lambda. There is a minimum value present, when 
lambda = 0.5. At this respective value, the RMSE is 0.8567. 

```{r RMSE Calculation}
# Define the RMSE Function
RMSE <- function(ratings_T, ratings_P){
  sqrt(mean((ratings_T - ratings_P)^2))
}

# Observing the RMSE based on lambdas value
lambdas <- seq(0,5,.5)
rmses <- sapply(lambdas, function(l){
  mu <- mean(edx$rating)
  
  b_i <- edx %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n() + l))
  
  b_u <- edx %>%
    left_join(b_i, by='movieId') %>% 
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n() +l))
  
  ratings_P <- edx %>%
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i +  b_u) %>% .$pred
  
  return(RMSE(ratings_P, edx$rating))
})

qplot(lambdas, rmses) + xlab("Lambdas") + ylab("RMSEs")
```

Interpreting the graph, the RMSE value is calculated below:

```{r Determining the RMSEs and Lambdas Value}
# Using the which.min() command to determine the lambdas value
print(c("The RMSEs value is:", round(min(rmses), 5)))
print(c("The lambdas (l) value is:", lambdas[which.min(rmses)]))
```

**After tuning with a new lambdas value, the new RMSE value becomes 0.8258.** 

```{r Final Calculation of RSME}
# Use the Model on the Validation Data for Confirmation
mu <- mean(validation$rating)
l <- 0.5
b_i <- validation %>%
  group_by(movieId) %>%
  summarize(b_i = sum(rating - mu)/(n() + l))

b_u <- validation %>%
  left_join(b_i, by='movieId') %>% 
  group_by(userId) %>%
  summarize(b_u = sum(rating - b_i - mu)/(n() +l))

ratings_P <- validation %>%
  left_join(b_i, by = "movieId") %>%
  left_join(b_u, by = "userId") %>%
  mutate(pred = mu + b_i +  b_u) %>% .$pred

RMSE(ratings_P, validation$rating)
```

# 4. Conclusion

The MovieLens data set provided countless data interpretations to the user. 
Significant discoveries include the popular genre selections, distribution
of the star ratings, and top movie titles. Additionally, discovering that the 
rating average remained relatively constant within the past decades has shown
that people from different generations tend to rate movies similarly. As a
result, it would be of best interest to emphasize areas such as genre, title 
of the movies. 

The RMSE value resulted in a value of 0.8258, a value that is close to 1, but 
not overtrained such that some potential errors may arise. 

There are current limitations based on the userId and the provided details, 
however; more work is needed to enhance the movie recommendation system. For
example, future work could include principal component analysis (PCA), a 
technique widely used to reduce the dimensionality of a large data set, and 
to observe any correlating information between different variables that may 
have been unnoticed in the earlier data interpretation stage. Using this 
powerful tool, other statistical analyses can be conducted to better suit 
the users' needs.

# 5. References
1. F. Maxwell Harper and Joseph A. Konstan. 2015. The MovieLens Datasets: 
History and Context. ACM Transactions on Interactive Intelligent Systems 
(TiiS) 5, 4, Article 19 (December 2015), 19 pages. DOI=http://dx.doi.org/10.1145/2827872




